\mbox{Verifying Imperative Higher-Order Programs}

### 2009-10-08 papers/esop2010/main.tex
go

The purpose of the present paper is to propose a language suited for reasoning
about effectful higher-order programs.  Higher-order functions are a key
feature, not only of purely functional languages like Haskell and the pure
fragment of ML, but also in combination with imperative features, like
references or arrays. However, reasoning about higher-order functions is
difficult, and especially so in the presence of effectful computations. Any
function's specification depends on its arguments. In a higher-order setting,
arguments may be functions that have their own specification. In consequence,
to be modular, the specification of a higher-order function can only be
formulated {\em depending on the specifications of its functional arguments}.
In a setting where side effects are possible, one also needs to formulate the
specification {\em depending on the side effects of the functional arguments}.
We claim that to be truly modular, the specification of a higher-order
function must be generic with respect to the specification {\em and} the
effect of its arguments.

@@

\paragraph{Outline.}
Section~\ref{sec:genpic} introduces our language using a motivating example.
Sections~\ref{sec:syntax} and \ref{sec:wp} give the details about the
programming language and the weakest precondition calculus.  In
section~\ref{sec:examples} we present our implementation and show some
examples. Finally, section~\ref{sec:related} discusses related work.

@@

Our programming language (Fig.~\ref{fig:syntax}) is basically the λ-calculus
with \letml-bindings, where there are two different λ-bindings. The first one
($λx:τ.t$) describes a pure, non-recursive binding, and the second one
($\recml~f~(x:τ).\{p\}t\{q\}$) describes recursive, potentially impure
functions; $p$ and $q$ in this term are {\em specifications}, which are
of the same syntax as programs, but are differently typed (typing will be
discussed in the next subsection). The term $p$ corresponds to the {\em
precondition}, $q$ to the {\em postcondition} of the function. The analogy to
Hoare triples is of course obvious. As usual, there are also variables,
\letml-bindings and an \ifml-\thenml-\elseml-construct. Variables on the type
level can be generalized at \letml-boundaries and have to be instantiated when
the corresponding program variable is used.

@@

In alias-free programs, the operations $\oplus$ and $|_ε$ on stores we
previously introduced can be executed {\em statically}, i.e. just by looking
at the domains of the stores. Also, we statically know which value (or which
named component) the accessor {\tt !!} will get from a store.

With the aliasing restriction, complex programs using shared mutable data
structures are impossible to prove. Programs with references in lists,
references in references and so on can be written, but only with the severe
limitation that a list can only contain a single reference (potentially
several times) and that a reference can only contain always the same
(statically known) reference. However, references still are first-class
values, i.e. they can be passed to functions, returned by functions and stored
in data structures, with the above limitations.

@@

Let's move on to the specification of {\tt ymemo}. We first define the {\tt
realizes} predicate between a pure function {\tt f0} and an impure one
modifying the table. This predicate states that {\tt f} {\em realizes}, i.e.
returns the same result as, {\tt f0} and additionally, if the table stored
valid key/value pairs, this is also the case after executing {\tt f}.
Actually, this is precisely the specification of {\tt memo}, so we could
easily prove {\tt realizes f0~(memo f0)} for any pure {\tt f0}. The definition
of the predicate uses the syntactic sugar for Hoare triples we introduced in
section~\ref{sec:genpic}.

>>>>

The purpose of the present paper is to propose a language suited for reasoning
about effectful higher-order programs.  Higher-order functions are a key
feature, not only of purely functional languages like Haskell and the pure
fragment of ML, but also in combination with imperative features, like
references or arrays. However, reasoning about higher-order functions is
difficult, and especially so in the presence of effectful computations. Any
function's specification depends on its arguments. In a higher-order setting,
arguments may be functions that have their own specification. As a consequence,
to be modular, the specification of a higher-order function can only be
formulated {\em depending on the specifications of its functional arguments}.
In a setting where side effects are possible, one also needs to formulate the
specification {\em depending on the side effects of the functional arguments}.
We claim that to be truly modular, the specification of a higher-order
function must be generic with respect to the specification {\em and} the
effect of its arguments.

@@

\paragraph{Outline.}
Section~\ref{sec:genpic} introduces our language using a motivating example.
Sections~\ref{sec:syntax} and \ref{sec:wp} give the details about the
programming language and the weakest precondition calculus.  In
section~\ref{sec:examples} we present our implementation and some
examples. Finally, section~\ref{sec:related} discusses related work.

@@

Our programming language (Fig.~\ref{fig:syntax}) is basically the λ-calculus
with \letml-bindings, where there are two different λ-bindings. The first one
($λx:τ.t$) describes a pure, non-recursive binding, and the second one
($\recml~f~(x:τ).\{p\}t\{q\}$) describes recursive, potentially impure
functions; $p$ and $q$ in this term are {\em specifications}, which are
of the same syntax as programs, but are differently typed (typing will be
discussed in the next subsection). The term $p$ corresponds to the {\em
precondition}, $q$ to the {\em postcondition} of the function. The analogy to
Hoare triples is obvious. As usual, there are also variables,
\letml-bindings and an \ifml-\thenml-\elseml-construct. Variables on the type
level can be generalized at \letml-boundaries and have to be instantiated when
the corresponding program variable is used.

@@

In alias-free programs, the operations $\oplus$ and $|_ε$ on stores can be
executed {\em statically}, i.e. just by looking at the domains of the stores.
Also, we statically know which value (or which named component) the accessor
{\tt !!} will get from a store.

With the aliasing restriction, complex programs using shared mutable data
structures are impossible to prove. Programs with references in lists,
references in references, etc., can be written, but only with the severe
limitation that a list can only contain a single reference (potentially
several times) and that a reference can only contain always the same
(statically known) reference. However, references still are first-class
values, i.e. they can be passed to functions, returned by functions and stored
in data structures, with the above limitations.

@@

Let us move on to the specification of {\tt ymemo}. We first define the {\tt
realizes} predicate between a pure function {\tt f0} and an impure one
modifying the table. This predicate states that {\tt f} {\em realizes}, i.e.
returns the same result as, {\tt f0} and additionally, if the table stored
valid key/value pairs, this is also the case after executing {\tt f}.
Actually, this is precisely the specification of {\tt memo}, so we could
easily prove {\tt realizes f0~(memo f0)} for any pure {\tt f0}. The definition
of the predicate uses the syntactic sugar for Hoare triples we introduced in
section~\ref{sec:genpic}.
